Liveness-Based Garbage Collection for Lazy Languages

### 2016-04-14 paper/cfg-to-nfa.tex
Updated NFA and DFAs

We circumvent the problem of  undecidability by over approximating the
CFG by non-deterministic finite state automata (NFA) using
%% The NFAs  are then simplified using the
%%$\hookrightarrow$ rules.   Finally the  simplified NFAs
%%are converted to DFAs.
%%
%% We use  the algorithm by 
Mohri   and   Nederhof~\cite{mohri00regular}   method.   This   method
transforms a  given CFG  into a strongly  regular grammar.

In our example, the rule  $\var{\Df{\length}{1}}$ has two context free
productions,           $\acdr\var{\Df{\length}{1}}\clazy$          and
$\clazy\var{\Df{\length}{1}}{\clazy}$.  We transform this grammar into
a strongly regular grammar in two steps,

\begin{enumerate}
\item
Add a  new non-terminal  $\var{\Df{\length}{1'}}$ to the  grammar with
the rule $\var{\Df{\length}{1'}} \rightarrow \epsilon$.
\item
Consider each production of $\var{\Df{\length}{1}}$,
\begin{enumerate}
\item
 replace     $\var{\Df{\length}{1}}     \rightarrow     \clazy$,     by
 $\var{\Df{\length}{1}} \rightarrow \clazy\var{\Df{\length}{1'}}$
\item
 replace
 $\var{\Df{\length}{1}} \rightarrow \clazy\var{\Df{\length}{1}}\clazy$,
 by  $\var{\Df{\length}{1}}  \rightarrow  \clazy\var{\Df{\length}{1}}$
 and $\var{\Df{\length}{1'}} \rightarrow \clazy\var{\Df{\length}{1'}}$
\item
 replace
 $\var{\Df{\length}{1}} \rightarrow \acdr\var{\Df{\length}{1}}\clazy$,
 by $\var{\Df{\length}{1}} \rightarrow \acdr\var{\Df{\length}{1}}$ and
 $\var{\Df{\length}{1'}} \rightarrow \clazy\var{\Df{\length}{1'}}$
\end{enumerate}
\end{enumerate}

The    resulting    grammar    fragment    for    %the    non-terminal
$\var{\Df{\length}{1}}$   after   the  Mohri-Nederhof   transformation
is:               \begin{eqnarray*}              \var{\Df{\length}{1}}
&\rightarrow& \clazy\var{\Df{\length}{1'}} \mid \acdr\var{\Df{\length}{1}} \mid \clazy\var{\Df{\length}{1}}\\ \var{\Df{\length}{1'}}
&\rightarrow& \clazy\var{\Df{\length}{1'}} \mid \epsilon \end{eqnarray*}
Notice that the language generated  by the strongly regular grammar is
a superset of the original CFG.

The strongly regular grammar is converted  into a set of NFAs, one for
each $\var{\Lanv{x}{}}$.  The  $\hookrightarrow$ simplification is now
done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
pairs  of consecutive  edges  labeled \bcar\acar\  or \bcdr\acdr\  and
constructing  the $\epsilon$-closure  till a  fixed point  is reached,
after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  The
details of the  algorithm, its correctness and  termination proofs are
given in~\cite{karkare07liveness,asati14lgc}.

@@

\begin{wrapfigure}{r}{0.25\textwidth}
\scalebox{0.5}{
          \begin{tabular}{cc}
          \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
          \begin{pspicture}(0,-3)(50,12)
            \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}} \putnode{t1}{t0}{15}{0}{\pscirclebox{\mbox{\ \ \ \
  }}} \hspace{5mm} \putnode{t2}{t1}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \
  }}} \psset{arrows=->} \ncline{t0}{t1} \ncline{t1}{t2} \putnode{l0}{t1}{7}{2}{\clazy} \nccurve[angleA=45,
  angleB=135,                                                 ncurv=4,
  nodesep=-1]{t1}{t1} \putnode{l1}{t1}{0}{8}{\acdr, \clazy} \nccurve[angleA=45,
  angleB=135,                                                 ncurv=3,
  nodesep=-1]{t2}{t2} \putnode{l2}{t2}{0}{8}{\clazy}
\end{pspicture}
&
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,-3)(30,12)
  \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}} \putnode{t1}{t0}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \
  }}}     \psset{arrows=->}    \ncline{t0}{t1}     \nccurve[angleA=45,
  angleB=135,                                                 ncurv=3,
  nodesep=-1]{t1}{t1} \putnode{l1}{t1}{0}{8}{\acdr}
\end{pspicture}
\\
(a) & (b)
 \end{tabular}}
\caption{(a) The  grammar rules for \var{\Df{\length}{1}}
  converted      into      an      automaton     and      (b)      its
  DFA.}\label{fig:example-automata}
\figrule
\vspace*{-5mm}
\end{wrapfigure}

Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
from the grammar for \var{\Df{\length}{1}}, and the final DFA is shown
in  Figure~\ref{fig:example-automata}(b).  This  expectedly says  that
for  a  demand  $\sigma_{\length}$,   the  liveness  of  the  argument
of \length\ is $\acdr^{*}$ (the spine of the list is traversed).

>>>>

We circumvent the problem of  undecidability by over approximating the
CFG by non-deterministic finite state automata (NFA) using
%% The NFAs  are then simplified using the
%%$\hookrightarrow$ rules.   Finally the  simplified NFAs
%%are converted to DFAs.
%%
%% We use  the algorithm by 
Mohri   and   Nederhof~\cite{mohri00regular}   method.   This   method
transforms a CFG $G$ into a strongly  regular grammar $R$ such that $\lang{G} \subseteq \lang{R}$. 
%%
If a CFG consists of a set of mutually recursive non-terminals such that the rules involved are not all left regular or not all right regular, then the method breaks the rules into right regular rules by introducing fresh non-terminals. For our example, the rule  $\var{\Df{\length}{1}}$ has two non regular productions,           $\acdr\var{\Df{\length}{1}}\clazy$          and
$\clazy\var{\Df{\length}{1}}{\clazy}$.  The steps for transforming these productions into
right regular productions are:

\begin{enumerate}
\item
Add a  new non-terminal  $\var{\Df{\length}{1'}}$ to the  grammar with
the rule $\var{\Df{\length}{1'}} \rightarrow \epsilon$.
\item
 Replace     $\var{\Df{\length}{1}}     \rightarrow     \clazy$,     by
 $\var{\Df{\length}{1}} \rightarrow \clazy\var{\Df{\length}{1'}}$
\item
 Replace
 $\var{\Df{\length}{1}} \rightarrow \clazy\var{\Df{\length}{1}}\clazy$,
 by  $\var{\Df{\length}{1}}  \rightarrow  \clazy\var{\Df{\length}{1}}$
 and $\var{\Df{\length}{1'}} \rightarrow \clazy\var{\Df{\length}{1'}}$
\item
  Replace
 $\var{\Df{\length}{1}} \rightarrow \acdr\var{\Df{\length}{1}}\clazy$,
 by $\var{\Df{\length}{1}} \rightarrow \acdr\var{\Df{\length}{1}}$ and
 $\var{\Df{\length}{1'}} \rightarrow \clazy\var{\Df{\length}{1'}}$
\end{enumerate}

Mohri and Nederhof~\cite{mohri00regular} provide the generalization of
these  steps.  The  rules for  
$\var{\Df{\length}{1}}$ after the transformation are:
\begin{eqnarray*}              
  \var{\Df{\length}{1}}   &\rightarrow&   \clazy\var{\Df{\length}{1'}}
  \mid                 \acdr\var{\Df{\length}{1}}                 \mid
  \clazy\var{\Df{\length}{1}}\\  \var{\Df{\length}{1'}}  &\rightarrow&
  \clazy\var{\Df{\length}{1'}} \mid \epsilon
\end{eqnarray*}

The strongly regular grammar is converted  into a set of NFAs, one for
each non-terminal.  The  $\hookrightarrow$ simplification is now
done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
pairs  of consecutive  edges  labeled \bcar\acar\  or \bcdr\acdr\  and
constructing  the $\epsilon$-closure  till a  fixed point  is reached,
after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  The
details of the  algorithm, its correctness and  termination proofs are
given in~\cite{karkare07liveness,asati14lgc}.

@@

\newcommand{\nfaD}{%
  \scalebox{.5}{\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
    \begin{pspicture}(0,-3)(50,12)
      \putnode{t0}{origin}{5}{0}{} 
      \putnode{t1}{t0}{15}{0}{\pscirclebox{\mbox{\ \ \ \      }}}
      \hspace{5mm} 
      \putnode{t2}{t1}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}} 
      \psset{arrows=->} \ncline{t0}{t1} 
      \ncline{t1}{t2} 
      \putnode{l0}{t1}{7}{2}{\clazy} \nccurve[angleA=45, angleB=135, 
        ncurv=4, nodesep=-1]{t1}{t1} 
      \putnode{l1}{t1}{0}{8}{\acdr, \clazy} \nccurve[angleA=45,
        angleB=135, ncurv=3, nodesep=-1]{t2}{t2} 
      \putnode{l2}{t2}{0}{8}{\clazy}
    \end{pspicture}
}}
\newcommand{\dfaD}{%
    \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
    \begin{pspicture}(0,-3)(30,12)
      \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}} \putnode{t1}{t0}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \
      }}}     \psset{arrows=->}    \ncline{t0}{t1}     \nccurve[angleA=45,
        angleB=135,                                                 ncurv=3,
        nodesep=-1]{t1}{t1} \putnode{l1}{t1}{0}{8}{\acdr}
    \end{pspicture}
}
The  NFA for \var{\Df{\length}{1}} is 
\nfaD\ and the final DFA is \dfaD.
This  expectedly says  that
for  a  demand  $\sigma_{\length}$,   the  liveness  of  the  argument
of \length\ is $\acdr^{*}$ (the spine of the list is traversed).
